 

### 2023-05-25 markdown/methods.Rmd
validation ellaborated in the methods

@@

##### Differential expression 
To test for differential expression, the analysis utilizes the "limma" package from R Bioconductor \citep{Phipson2016}. This package employs a linear model approach to determine the fold expression between sample groups. Before initiating the analysis, two matrices need to be computed. The design matrix identifies samples based on their sample type and defines the experimental design. An algorithm within the program's backend logic is utilized to automatically generate this matrix according to the user's selection in the interface. The expression matrix contains intensities for each identified protein and is obtained at the end of the pipeline. By calling a function with the design and expression matrices as arguments, the contrast matrix is calculated, enabling the user to decide which comparisons to consider and also perform tests on multiple factors.

The subsequent step involves creating a linear model between groups using log-ratios of their expression values. In a two-way design, the expression of the gene $y$ can be explained as $Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \epsilon$, where $\beta_0$ represents the intercept, which can be interpreted as the mean expression of the gene. $\beta_1$ represents difference in mean of the treatment (or condition) on the discrete or continuos variable $X_{1}$, and $\epsilon$ acts as an error term. The interface allows for additive models with multiple factors, allowing the user to add factors according to various scenarios. For example, in a two-factor model, it would look like: $Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + \epsilon$. The principle remains the same for models with multiple factors. When visualizing the linear model, the x-axis would represent the gene for the sample types or other factors, whereas the y-axis would account for the expression value. By drawing a line through the cloud of data points the model is constructed by minimizing the quadratic distance to every data-point, called the least-square method or also ordinary least-square method. With the regression line the above mentioned equation can be constructed.

@@

When observing multiple batches at once, as already explained in the chapter batch correction this effect needs to be taken into consideration when performing differential expression analysis. This is done by the multifactor option in the user interface by selecting the batch as a co factor and inclusion in the design matrix.

>>>>

#### Validation of the methods applied to the dataset
In the context of the validation of methods applied to a dataset, the mutual information calculation is used to assess the impact of computations on the dataset and the information it carries. 
Mutual information, denoted as $I(X;Y)$, is a measure of the dependence or information shared between two variables, $X$ and $Y$, in this case sample types. It is calculated using the equation $I(X;Y) = H(X,Y) - H(X|Y) - H(Y|X)$, where $H$ represents the entropy, $H(X|Y)$ and $H(Y|X)$ denote for the both conditional probabilities and $H(X,Y)$ for the joint entropy. 
Entropy, as a concept in information theory, quantifies the level of disorder or uncertainty associated with a random event. Events that are highly likely to occur are considered less informative than events that are more unexpected or less likely. In a Venn diagram $I(X;Y)$ (the mutual information) corresponds to the intersection of the two conditional entropys $H(X|Y)$ and $H(Y|X)$. To calculate the mutual information between two variables $X$ and $Y$ we can alternatively express the above equation as $I(X;Y) = D_{KL} (P_{(X,Y)} || P_{X} \otimes P_{Y})$. $D_{KL}$ denotes for the Kullback-Leibler divergence which is a concept of relative entropy between two variables and tells us how much 2 probability distributions differ from each other and $P_{X} \otimes P_{Y}$ represents the product of the marginals. The underlying principle of divergence is the integral asymmetry in Bayesian inference, which means that divergence does not satisfy the triangle inequality, and differences are not the same depending on the starting point. By comparing the mutual information before performing any calculations with the mutual information calculated after the computations, it is possible to assess whether the applied methods have significantly altered the dependence or information content. The goal is to choose a method which preserves the information contributed by each biological sample type to the experiment.
The R package "infotheo" is utilized to perform the estimates for the entropy of the two variables. Mutual information is returned as natural unit of information (=nat), a measurement proportional to the to the Shannon entropy ($1nat = \frac{1}{ln2} shannons$). Before calculating the pairwise mutual information of the expression matrix, values were discretized into bins with the discretize function of above package.
The mutual information is calculated within each sample type, and it can also be calculated between a selection of sample types to determine if the computations performed affect the information content. After that all pairwise mutual information within the expression set before and after the pipeline is obtained. The pairwise differences can be observed as a boxplot for a particular sample type indicating the change in dependence within and also compared to another sample type.

@@

##### Differential expression 
To test for differential expression, the analysis utilizes the "limma" package from R Bioconductor \citep{Phipson2016}. This package employs a linear model approach to determine the fold expression between sample groups. Before initiating the analysis, two matrices need to be computed. The design matrix identifies samples based on their sample type and defines the experimental design. An algorithm within the program's backend logic is utilized to automatically generate this matrix according to the user's selection in the interface. The expression matrix contains intensities for each identified protein and is obtained at the end of the pipeline. By calling a function with the design and expression matrices as arguments, the contrast matrix is calculated, enabling the user to decide which comparisons to consider and also perform tests on multiple factors. When observing multiple batches at once, as already explained in the chapter batch correction this effect needs to be taken into consideration when performing differential expression analysis. This is done by the multi factor option in the user interface by selecting the batch as a co factor and inclusion in the design matrix. The advantage of this procedure is, that we can quantify the biological relevance of working in multiple batches and avoid publishing false positive results.
The multi factor option uses an additive model as a design matrix, whereas continuous variables (such as size, weight etc.) can also be used to build the statistical model.

The subsequent step involves creating a linear model between groups using log-ratios of their expression values. In a two-way design, the expression of the gene $y$ can be explained as $Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \epsilon$, where $\beta_0$ represents the intercept, which can be interpreted as the mean expression of the gene. $\beta_1$ represents difference in mean of the treatment (or condition) on the discrete or continuous variable $X_{1}$, and $\epsilon$ acts as an error term. The interface allows for additive models with multiple factors, allowing the user to add factors according to various scenarios. For example, in a two-factor model, it would look like: $Exp_{Y} = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + \epsilon$. The principle remains the same for models with multiple factors. When visualizing the linear model, the x-axis would represent the gene for the sample types or other factors, whereas the y-axis would account for the expression value. By drawing a line through the cloud of data points the model is constructed by minimizing the quadratic distance to every data-point, called the least-square method or also ordinary least-square method. With the regression line the above mentioned equation can be constructed.

@@

Limma outputs a table with all proteins, their respective logfold change, p-values and adjusted p-values. The user can choose different methods for the correction of the p-values, which is necessary when testing the same hypothesis multiple times. A common method in omics studies is the false discovery rate (=fdr) by Benjamini Hochberg \citep{Benjamini1995}. By ranking all the p-values from smallest to largest the p-values are adjusted sequentially using the formula $p_(i) = \frac{r_i}{m} Q$. The rank in the p-value table is denoted as $r_i$   , $m$ reflects the total number of tests and $Q$ represents false discovery rate which is set to 5%. The second option to choose is the Benjamini & Yekutieli (=BY) \citep{Benjamini2001}method, which is a further development of the fdr by the same statistician and has a wider range of application based on the dependencies for the test.
The third and most conservative option is the Holm method \citep{Holm1979} which controls the error sequentally in a family wise manner, which is considered as traditional in regards to the above mentioned experimental wise error rates. The above mentioned correction methods are ordered in the way from the most liberal to the most conservative one. The selection is supposed to give the user the possibility questioning results on multiple levels.
